The Evolution of Phenotypically Invariant Genetic Networks

### 2018-08-16 review-responses.tex

\begin{quote}
I have received the evaluations of two reviewers, and I have read the paper
myself. First, sorry for the long time it took to review your manuscript; it
was reviewed on the timescale of mathematicians (for Reviewer2), but both
reviewers provided thorough evaluations of your work (which in my opinion is
better than a quick but superficial review).

Your manuscript is quite long already, and the reviewers' suggestions of
modifications and clarification, which need to be implemented, will make the
manuscript even longer. I however share the reviewers' opinion that the
manuscript's two parts are only loosely related. My suggestion is therefore to
publish them separately. Although connecting the systems biology and the
quantitative genetics parts is indeed an exciting endeavor, this is not really
achieved in the current version of the manuscript, and it would be more
profitable to first better describe each part separately. Regarding the
connexion between the two parts, please also pay particular attention to R2's
first specific comment about dimensionality.

This may be because I am more familiar with Fisher's geometric model than
systems biology models, but it seems to me that the first part is more novel,
and should therefore deserve your attention first, should you follow my
suggestion of publishing the two parts separately.
\end{quote}

\begin{point}{Format:}
    Please add line numbers to your manuscript; it is straightforward
with the lineno package in LaTeX.  Please also do not increase text width too
much, as this decreases legibility (LaTeX's default settings already optimize
the number of characters per line).  Finally, be careful about not confusing
citep and citet (a lot of missing parentheses around citations in the second part).
\end{point}

@@

\begin{point}{Example 1:}
    ``and so its phenotype'' whose phenotype? (the subject of the sentence is ``the input'')
\end{point}

\reply{
    Fixed. \revref
}

\begin{point}{p5, Note on implementation}
    ``the closure of span(B)'' may not be
    understandable by most of Evolution's readers; please explain what this means.
\end{point}

\reply{
    We've added a clarifying note \revref
    but we're hoping that readers not familiar with linear algebra will realize
    that they can skip this ``note on implementation'' without loss of continuity.
}

@@

\begin{point}{Figures 1-3:}
    In figures 1 and 3, there are two types of arrow heads (+ and -
effects), but not in Figure 2, probably because the sign of the effect depends
on the value of $\tau$. Maybe use a  different type of arrow heads for
interactions that may change signs, to avoid the confusion with positive
effects?
\end{point}

\reply{
}

\begin{point}{p7, Hybrid incompatibility:}
    Please provide a rationale for choosing a particular weight $\rho$.
\end{point}

\reply{
    Good suggestion; added. \revref
}

\begin{point}{p11, 2nd paragraph}
    Try to better relate these different studies to yours
(e.g., what you add, what they do and you do not). In addition to R1's
suggestions, consider \citet{weinreich2013fishers} and \citet{blanquart2016epistasis}.
% Weinreich, D. M. and Knies, J. L. (2013), FISHER'S GEOMETRIC MODEL OF ADAPTATION MEETS THE FUNCTIONAL SYNTHESIS: DATA ON PAIRWISE EPISTASIS FOR FITNESS YIELDS INSIGHTS INTO THE SHAPE AND SIZE OF PHENOTYPE SPACE. Evolution, 67: 2957-2972. doi:10.1111/evo.12156
% François Blanquart and Thomas Bataillon Epistasis and the Structure of Fitness Landscapes: Are Experimental Fitness Landscapes Compatible with Fisher’s Geometric Model?  GENETICS Early online April 6, 2016
\end{point}

\reply{
    This part has been removed.
}

@@

\reply{
}

\begin{point}{Figure 2:}
    Colors are hardly visible
\end{point}

\reply{
}

\begin{quote}
    \textbf{(II)} The popgen part. This part models the ``system drift'', which is nothing else
but neutral drift of a population along a high-fitness ridge in an epistatic
landscape (conceptual figure 4), right? If two populations in allopatry drift
in different directions, this can lead to hybrid incompatibilities, which are
uncovered upon secondary contact. Fitness is modeled by the weighted distance
from the optimal impulse response function (eq 5). This is a natural
assumption.
\end{quote}

\reply{
    \plr{I'm leaving these points because they might apply to whatever handwaving 
    we replace the quant gen stuff with.}
}

@@

\begin{point}{p7: fig 4.}
    ``The distribution of $F_2$ ... homozygotes''. I do not really grasp the meaning of this
sentence. More importantly, I find the purple cloud of dots ($F_2$ population) quite confusing.
It seems to me that the purple populations should be a cloud of points concentrated around
a point sitting on the red dotted line, i.e., a Gaussian distribution with the mean at the
average of the two parental populations. According to the authors, the distribution is bi-
modal with a peak below the optimal set (as displayed on the figure) and another distinct
peak sitting on the other side of the optimal. In fact, if I understand the computations of
the second part correctly, the red and purple means should coincide, but the red variance
should simply be greater.
\end{point}

\reply{
}

\begin{point}{p7: end of the page.}
    Why do you need $\rho$ to be square integrable?
\end{point}

\reply{
    Good point -- clearly some assumptions jointly on the long-time behavior of $h(t)$ and $\rho$ 
    are required for $D$ to be finite, but just saying that $\rho$ is square integrable doesn't suffice;
    we've removed the caveat. 
    \revref
    (It is always well-defined, anyhow, since the integrand is nonnegative.)
}

@@

\begin{point}{p9. Fig 5.} 
    In the left panel, it seems to me that the two parents are homozygotes since there
is a single $F_1$ possible offspring (dashed blue curve). It would be worth being more explicit.
For the right panels, could you be more explicit on the number of curves. If I understand
correctly, there are 16 possibilities due to recombination (2 per entries of the matrix) since
$F_1$ individuals are heterozygotes. Is that correct?
Finally, the labels on the y axis are not easy to read.
\end{point}

\reply{
}

\begin{point}{p11. Paragraph system drift.}
    ``move a random distance $\sigma$''. What is $\sigma$? I think it should
be $\sigma_N$ (the std deviation in the direction of the evolutionary ridge) to be consistent with
the assumption that the population drifts along the optimal set. I believe that this is what
is assumed thereafter. Also, the sentence ``It therefore seems .... as cloud of points of width
$\sigma$'' is not very accurate, since the covariance matrix is not the identity.
\end{point}

\reply{
    This part has been removed.
}

\reply{
    This part has been removed.
}

\reply{
    This part has been removed.
}

\reply{
    This part has been removed.
}

\reply{
    This part has been removed.
}

\reply{
    This part has been removed.
}

\begin{point}{}
    Finally, it would be worth mentioning several old and recent works relating genetic drift
to speciation: Yamagushi and Iwasa, several articles by Gavrilets et al. (I think several
citations are missing here), and Mirò Pina and Schertzer.
\end{point}

### 2018-08-16 kalman_walk_paper.tex
started on responses

We use a model of gene regulatory networks
that describes the temporal dynamics of
a collection of $n$ coregulating molecules
-- such as transcription factors -- as well as external or environmental inputs.
% We model the temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
% within an organism, that may also be affected by temporally varying signals from the environment.
%% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
%%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
%%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
The vector of $m$ ``inputs'' determined exogenously to the system is denoted $u(t)$,
and the vector of $\ell$ ``outputs'' is denoted $\phi(t)$.
The output is merely a linear function of the internal state:
$\phi_i(t) = \sum_j C_{ij} \kappa_j(t)$
for some matrix $C$.
Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype}
(meaning the ``visible'' aspects of the organism),
and in contrast refer to $\kappa$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
Although $\phi$ may depend on all entries of $\kappa$,
it is usually of lower dimension than $\kappa$,
and we tend to think of it as the subset of molecules relevant for survival.
The dynamics are determined by
the matrix of regulatory coefficients, $A$,
a time-varying vector of inputs $u(t)$,
and a matrix $B$ that encodes the effect of each entry of $u$ on the elements of the kryptotype.
%Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regulatory network), given by the triple $(A,B,C)$,
%and by its environment $u(t)$.
The rate at which the $i^\text{th}$ concentration changes
is a weighted sum of the concentrations
as well as the input:
\begin{equation}\label{eqn:system}
   \begin{aligned}
    \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\
    \phi(t) &= C \kappa(t) .
  \end{aligned} 
\end{equation}
Furthermore, we always assume that $\kappa(0) = 0$,
so that the kryptotype measures deviations from initial concentrations. 
Here $A$ can be any $n \times n$ matrix, $B$ any $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
with usually $\ell$ and $m$ less than $n$.
We think of the system as the triple $(A,B,C)$,
which translates (time-varying) $m$-dimensional input $u(t)$ into the $\ell$-dimensional output $\phi(t)$.
Under quite general assumptions,
we can write the phenotype as
  \begin{align}
    \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
  \end{align}
which is a convolution of the input $u(t)$ with the system's \emph{impulse response},
which we denote as $h(t) := Ce^{A t}B$.

@@

\begin{example}[An oscillator]\label{ex:oscillator}
For illustration, we consider an extremely simplified model of oscillating gene transcription,
as for instance is found in cell cycle control or the circadian rhythm.
% Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}.
There are two genes, 
whose transcript concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$, 
and gene-2 upregulates gene-1, while gene-1 downregulates gene-2 with equal strength.
Only the dynamics of gene-1 are consequential to the oscillator 
(perhaps the amount of gene-1 activates another downstream gene network). 
Lastly, both genes are equally upregulated by an exogenous signal.
The dynamics of the system are described by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
        \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t) \\
        \phi(t) &= \kappa_{1}(t)   .
    \end{align*}
In matrix form the system regulatory coefficients are given as,
$A \!=\! \left[\begin{smallmatrix} 0 & 1 \\ -1 & 0 \end{smallmatrix}\right]$, 
$B \! =\! \left[\begin{smallmatrix} 1 \\ 1 \end{smallmatrix}\right]$,
and $C \!=\! \left[\begin{smallmatrix} 1 & 0 \end{smallmatrix}\right]$.
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}
Suppose the input is an impulse at time zero (a delta function), and so its phenotype is equal to its impulse response,
\begin{align*}
\phi(t) = h(t) = \sin t + \cos t  .
\end{align*}
The system and its dynamics are referred to in Figure \ref{fig:oscillator}. We return to the evolution of such a system below.   
\end{example}

@@

\emph{Note on implementation:}
The \emph{reachable subspace},
which we denote by $\reachable$,
is defined to be the closure of $\spn(B)$ under applying $A$,
% (or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
% Analogously to this, we define
and the \emph{unobservable subspace}, 
%$\mathcal{O}$, is the closure of $\spn(C^T)$ under applying $A$.
denoted $\unobservable$, is the largest $A$-invariant subspace
contained in the null space of $C$.
% and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
The four subspaces, $\rno$, $\ro$, $\nrno$, and $\nro$
are defined from these by intersections and orthogonal complements --
$\ro$ refers to the both \emph{reachable and observable} subspace,
while $\nrno$ refers to the \emph{unreachable and unobservable} subspace,
and similarly for $\nro$ and $\rno$.

@@

\subsection*{Hybrid incompatibility}
Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
if the parents have different underlying systems.
%This leads to the question: 
How quickly do hybrid phenotypes break down as genetic distance between parents increases?
We will quantify how far a system's phenotype is from optimal
using a weighted difference between impulse response functions.
Suppose that $\rho(t)$ is a nonnegative, smooth, square-integrable weighting function,
$h_0(t)$ is the \emph{optimal} impulse response function
and define the ``distance to optimum'' of another impulse response function
to be
\begin{align}
\label{eqn:distance}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
Consider reproduction between a parent with system $(A, B, C)$ 
and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
We assume both are ``perfectly adapted'' systems, 
i.e., having impulse response function $h_0(t)$,
and their offspring has impulse response function $h_\epsilon(t)$.
A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ % is explicitly worked out in Appendix \ref{apx:H_calc}, and
shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
while $F_2$ hybrids are at distance proportional to $\epsilon$.
This is because an $F_1$ hybrid has one copy of each parental system,
and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
the parents both lie in $\allS$, which is the valley defined by $D$,
and so their midpoint only differs from optimal due to curvature of $\allS$.
In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
and homozygous for the other parental type in others;
this means that each coefficient of an $F_2$ may be equal to either one of the parents,
or intermediate between the two;
this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
as the distance between the parents.
The precise rate at which the phenotype of a hybrid diverges depends on the geometry
of the optimal set $\allS$ relative to segregating genetic variation.
% in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set) with respect to the coordinates.

>>>>

We use a model of gene regulatory networks
that describes the temporal dynamics of
a collection of $n$ coregulating molecules
-- such as transcription factors -- as well as external or environmental inputs.
% We model the temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
% within an organism, that may also be affected by temporally varying signals from the environment.
%% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
%%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
%%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
The vector of $m$ ``inputs'' determined exogenously to the system is denoted $u(t)$,
and the vector of $\ell$ ``outputs'' is denoted $\phi(t)$.
The output is merely a linear function of the internal state:
$\phi_i(t) = \sum_j C_{ij} \kappa_j(t)$
for some matrix $C$.
Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype}
(meaning the ``visible'' aspects of the organism),
and in contrast refer to $\kappa$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
Although $\phi$ may depend on all entries of $\kappa$,
it is usually of lower dimension than $\kappa$,
and we tend to think of it as the subset of molecules relevant for survival.
The dynamics are determined by
the matrix of regulatory coefficients, $A$,
a time-varying vector of inputs $u(t)$,
and a matrix $B$ that encodes the effect of each entry of $u$ on the elements of the kryptotype.
%Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regulatory network), given by the triple $(A,B,C)$,
%and by its environment $u(t)$.
The rate at which the $i^\text{th}$ concentration changes
is a weighted sum of the concentrations
as well as the input:
\begin{equation}\label{eqn:system}
   \begin{aligned}
    \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\
    \phi(t) &= C \kappa(t) .
  \end{aligned} 
\end{equation}
Furthermore, we always assume that $\kappa(0) = 0$,
so that the kryptotype measures deviations from initial concentrations. 
Here $A$ can be any $n \times n$ matrix, $B$ any $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
with usually $\ell$ and $m$ less than $n$.
We think of the system as the triple $(A,B,C)$,
which translates (time-varying) $m$-dimensional input $u(t)$ into the $\ell$-dimensional output $\phi(t)$.
Under quite general assumptions,
we can write the phenotype as \revpoint{AE}{4}
  \begin{align}
    \phi(t) = \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
  \end{align}
which is a convolution of the input $u(t)$ with the system's \emph{impulse response},
which we denote as $h(t) := Ce^{A t}B$.

@@

\begin{example}[An oscillator]\label{ex:oscillator}
For illustration, we consider an extremely simplified model of oscillating gene transcription,
as for instance is found in cell cycle control or the circadian rhythm.
% Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}.
There are two genes, 
whose transcript concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$, 
and gene-2 upregulates gene-1, while gene-1 downregulates gene-2 with equal strength.
Only the dynamics of gene-1 are consequential to the oscillator 
(perhaps the amount of gene-1 activates another downstream gene network). 
Lastly, both genes are equally upregulated by an exogenous signal.
The dynamics of the system are described by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
        \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t) \\
        \phi(t) &= \kappa_{1}(t)   .
    \end{align*}
In matrix form the system regulatory coefficients are given as,
$A \!=\! \left[\begin{smallmatrix} 0 & 1 \\ -1 & 0 \end{smallmatrix}\right]$, 
$B \! =\! \left[\begin{smallmatrix} 1 \\ 1 \end{smallmatrix}\right]$,
and $C \!=\! \left[\begin{smallmatrix} 1 & 0 \end{smallmatrix}\right]$.
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}
If the input is an impulse at time zero (a delta function), 
then the phenotype is equal to the impulse response: \revpoint{AE}{5}
\begin{align*}
\phi(t) = h(t) = \sin t + \cos t  .
\end{align*}
The system and its dynamics are referred to in Figure \ref{fig:oscillator}. We return to the evolution of such a system below.   
\end{example}

@@

\emph{Note on implementation:}
The \emph{reachable subspace},
which we denote by $\reachable$,
is defined to be the closure of $\spn(B)$ under applying $A$
(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$), \revpoint{AE}{6}
% Analogously to this, we define
and the \emph{unobservable subspace}, 
%$\mathcal{O}$, is the closure of $\spn(C^T)$ under applying $A$.
denoted $\unobservable$, is the largest $A$-invariant subspace
contained in the null space of $C$.
% and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
The four subspaces, $\rno$, $\ro$, $\nrno$, and $\nro$
are defined from these by intersections and orthogonal complements --
$\ro$ refers to the both \emph{reachable and observable} subspace,
while $\nrno$ refers to the \emph{unreachable and unobservable} subspace,
and similarly for $\nro$ and $\rno$.

@@

\subsection*{Hybrid incompatibility}
Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
if the parents have different underlying systems.
How quickly do hybrid phenotypes break down as genetic distance between parents increases?
We will quantify how far a system's phenotype is from optimal
using a weighted difference between impulse response functions.
Suppose that $\rho(t)$ is a nonnegative weighting function, \revpoint{2}{3}
$h_0(t)$ is the \emph{optimal} impulse response function
and define the ``distance to optimum'' of another impulse response function
to be
\begin{align}
\label{eqn:distance}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
In practice, we take $\rho(t) = \exp(-t/4\pi)$,
so that fitness is determined by the dynamics of the system over a few multiples of $2 \pi$,
but not longer. \revpoint{AE}{8}
Consider reproduction between a parent with system $(A, B, C)$ 
and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
We assume both are ``perfectly adapted'' systems, 
i.e., having impulse response function $h_0(t)$,
and their offspring has impulse response function $h_\epsilon(t)$.
A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ % is explicitly worked out in Appendix \ref{apx:H_calc}, and
shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
while $F_2$ hybrids are at distance proportional to $\epsilon$.
This is because an $F_1$ hybrid has one copy of each parental system,
and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
the parents both lie in $\allS$, which is the valley defined by $D$,
and so their midpoint only differs from optimal due to curvature of $\allS$.
In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
and homozygous for the other parental type in others;
this means that each coefficient of an $F_2$ may be equal to either one of the parents,
or intermediate between the two;
this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
as the distance between the parents.
The precise rate at which the phenotype of a hybrid diverges depends on the geometry
of the optimal set $\allS$ relative to segregating genetic variation.
% in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set) with respect to the coordinates.
